---
title: "Final Review - Citywide Payroll Data Analysis"
output:
  html_document:
    fig_height: 4
    highlight: pygments
    theme: spacelab
  pdf_document: default
---
### Reg. No: 18BCE1153

### Name: Ashwin Arun

####  Setup: Rstudio v1.2.5




## What problem are you trying to solve?

* * *
Monitor and compare the payment and benefit provision details of various job titles and departments across the city and to determine if the jobs provide sufficient basic pay and compensations with respect to the basic cost of living in LA.

* * *

## What data have you chosen?(Chosen Dataset, Source of dataset, Description of dataset, basic commands to describe dataset)

* * *
Dataset Name: Los Angeles City Payroll Data

Source of Dataset: Kaggle

Description of Dataset: 

This dataset is based on Los Angeles City and contains the data collected over the years from 2013-2016. It contains payroll information about every employee title in Los Angeles city departments. The database includes their entitled salary, monthly (or quarterly) payments, annual bonuses, various types of benefits and compensation amounts. 

* * *

## Frame your objectives
* * *
1.To identify the civil job or department that has positions offering below median salary. i.e. Insufficient salary
2.To plot/illustrate the variance of salary between the lowest and highest level of every department.
3.To outline the benefit provisions offered by various departments.
3.To derive a relationship between final quarterly payment and base pay of a particular department and predict the base pay of its employees.
* * * 
## Basic Operations on Dataset


#### Dimensions (rows & columns) of dataset
```{r}
data <- read.csv("data.csv")
dim(data)
```

#### Retrieving the various attribute/column names of dataset
```{r}
data <- read.csv("data.csv")
names(data)
```


#### Retrieving top 4 records of database
```{r}

data <- read.csv("data.csv")
head(data,4)
```

```{r}
data <- read.csv("data.csv")
str(data)
```
### Data Cleaning and Preparation
#### Creating a new dataset after cleaning the original dataset
```{r}
library(dplyr)
library(magrittr)
data <- read.csv("data.csv")


#Convert to lowercase
data[c(3,6,7,26,34)]<-lapply(data[c(3,6,7,26,34)], tolower)
data[26]<-lapply(data[26], tolower)
#remove extra white spaces
data[c(3,6,7,26,34)]<- lapply(data[c(3,6,7,26,34)], function (x) gsub("^\\s+|\\s+$", "",x))
#Converting num values
data[8:14] <- lapply(data[8:14], function(x) as.numeric(gsub("\\$","",x)))
data[16:24] <- lapply(data[16:24], function(x) as.numeric(gsub("\\$","",x)))
data[30:33] <- lapply(data[30:33], function(x) as.numeric(gsub("\\$","",x)))
#Substituting na,N/A values with NA
data<-data %>% mutate(Payroll.Department=replace(Payroll.Department,Payroll.Department=="N/A",NA)) %>%
  mutate(Benefits.Plan=replace(Benefits.Plan,Benefits.Plan=="na",NA))
#Removing rows with incorrect/corrupt info
data2<-data[!(data$Total.Payments<1),]
data2<-data2[!(data2$Projected.Annual.Salary==0),]
#Changing NA values to 0
data2$Hourly.or.Event.Rate[is.na(data2$Hourly.or.Event.Rate)] <- 0
data2$Lump.Sum.Pay[is.na(data2$Lump.Sum.Pay)] <- 0
data2$Overtime.Pay[is.na(data2$Overtime.Pay)] <- 0
#Removing MOU and MOU.Title attribute
data2 <- data2[-c(24,25,29)]

write.csv(data2,"C:\\Users\\Ashwin A\\Desktop\\VIT\\FALL SEM 2020-21\\DATA ANALYTICS\\DA PROJECT\\data_clean.csv",row.names=FALSE)

```
#### 1. BoxPlotting the Base Salary via Department
```{r}
#Boxplotting Base Salary 

library(ggplot2)
datac <- read.csv("C:\\Users\\Ashwin A\\Desktop\\VIT\\FALL SEM 2020-21\\DATA ANALYTICS\\DA PROJECT\\data_clean.csv")

ggplot(datac, aes(factor(`Department.Title`), Base.Pay/1000)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + xlab("Department Name") + ylab("Average Annual Salary")

```
#### Description:
* * *
This boxplot graph is meant to provide an overall idea about the median salary and its outliers of each department name. The average annual salary is considered in this graph so that the salary amount is unaffected by other factors such as benefits or overtime pay. As we can observe a few departments such as Airport and Police(LAPD) have a large number of positive outliers indicating a large pay gap between the top ranking officials with the average salary of other employees in the same department. 


#### 2. To identify the civil job/ departmental positions with below median salary.
```{r}
library(dplyr)
library(magrittr)
library(ggplot2)
library(png)

datac <- read.csv("C:\\Users\\Ashwin A\\Desktop\\VIT\\FALL SEM 2020-21\\DATA ANALYTICS\\DA PROJECT\\data_clean.csv")
par(mar=c(4,11,4,4))

#Creating a sub dataframe for jobs providing salary less than median salary
#2013
df1 <- datac %>% filter(Projected.Annual.Salary<58875,Year==2013,Employment.Type=="full time") %>% select(Year,Projected.Annual.Salary,Job.Class.Title) 
df1<-df1 %>% distinct(Job.Class.Title, .keep_all = TRUE) %>% arrange(Projected.Annual.Salary)
df1<-head(df1,10)
#2014
df2 <- datac %>% filter(Projected.Annual.Salary<58875,Year==2014,Employment.Type=="full time") %>% select(Year,Projected.Annual.Salary,Job.Class.Title)
df2<-df2 %>% distinct(Job.Class.Title, .keep_all = TRUE) %>% arrange(Projected.Annual.Salary)
df2<-head(df2,10)
#2015
df3 <- datac %>% filter(Projected.Annual.Salary<62676,Year==2015,Employment.Type=="full time") %>% select(Year,Projected.Annual.Salary,Job.Class.Title)
df3<-df3 %>% distinct(Job.Class.Title, .keep_all = TRUE) %>% arrange(Projected.Annual.Salary)
df3<-head(df3,10)
#2016
df4 <- datac %>% filter(Projected.Annual.Salary<64185,Year==2016,Employment.Type=="full time") %>% select(Year,Projected.Annual.Salary,Job.Class.Title)
df4<-df4 %>% distinct(Job.Class.Title, .keep_all = TRUE) %>% arrange(Projected.Annual.Salary)
df4<-head(df4,10)
#Barplotting 
par(mfrow=c(2,2))
barplot(main="Low Paying Occupations 2013",height=df1$Projected.Annual.Salary, names=df1$Job.Class.Title, col="#ff6600", horiz=T , las=2)
barplot(main="Low Paying Occupations 2014",height=df2$Projected.Annual.Salary, names=df2$Job.Class.Title, col="#ff6600", horiz=T , las=2)
barplot(main="Low Paying Occupations 2015",height=df3$Projected.Annual.Salary, names=df3$Job.Class.Title, col="#ff6600", horiz=T , las=2)
barplot(main="Low Paying Occupations 2016",height=df4$Projected.Annual.Salary, names=df4$Job.Class.Title, col="#ff6600", horiz=T , las=2)

```

#### Description:
* * *
This collection of graph depicts the low-paying occupations or positions classified by each year. The parameter (i.e. limit) is the average salary a person should earn to afford a living in the city of LA. In this result we have considered only the full-time positions and their projected annual salary to keep it unaffected by any variables like overtime, bonus payment etc. The objective of this visualisation is to provide job seekers an insight about the outliers.


#### 3. To plot/illustrate the variance of salary between the lowest and highest level of every department.
```{r}
library(dplyr)
library(magrittr)
library(ggplot2)
datac <- read.csv("C:\\Users\\Ashwin A\\Desktop\\VIT\\FALL SEM 2020-21\\DATA ANALYTICS\\DA PROJECT\\data_clean.csv")

#Generating max,min and diff by dept AND plotting using lollipop plot
df1diff<-datac %>% filter(Year==2013,Employment.Type=="full time") %>% group_by(Department.Title) %>% summarise(max_basepay = max(Base.Pay, na.rm=TRUE), min_basepay = min(Base.Pay, na.rm=TRUE)) %>% mutate(diff=(max_basepay-min_basepay)/1000) %>% arrange(diff)
p1<-ggplot(df1diff, aes(x=Department.Title, y=diff)) +coord_flip() + geom_point(size=3, alpha=0.6) + geom_segment(aes(x=Department.Title, xend = Department.Title, y=0, yend = diff)) + ggtitle("How much is the difference of salaries in 2013?")
p1

```
#### Description:
* * *
In this plot we are depicting the difference in the paygrade of each department in the year 2015. This certain type of plot is called a 'lollipop plot'. The difference between the highest paying and lowest paying position is measured in a scale by dividing the difference amount by 1000 and represented by the height of the plot. The point depicts the difference rate for each department.
As observed, large differences are noticable in some of the prominent departments such as police, water/power and airports. Such cases of extreme wage gap can be unhealthy for the overall economy apart from affecting the lower ranking employees.



#### 4.To identify the department that provides the most benefit package
```{r}
library(ggplot2)
library(reshape2)
library(dplyr)

benefit <- data %>% group_by(Benefits.Plan) %>% summarise(meanh=mean(Average.Health.Cost), meand=mean(Average.Dental.Cost),meanb=mean(Average.Basic.Life)) 
benefit <- benefit[-1,]
cmatrix_melted <- melt(benefit)
par(las=2)
ggplot(cmatrix_melted, aes(fill=variable, y=value, x=Benefits.Plan)) + geom_bar(position="stack", stat="identity")

```

#### Description:
* * *
This stacked barplot basically provides an idea about the distribution of various benefits packages by various benefit providers. The stacks indicate the package type. The major providers include police and fire. Rest of the employees are under city payroll department. From the graph we find that city offers the highest average basic life benefit while police department offers the most health benefit amount.



#### 5. To show the relationship between base pay, extra pay, overtime pay and average benefit cost via correlation and represent it using a heatmap
```{r}
library(ggplot2)
library(reshape2)
library(dplyr)

datasub <- select(datac,c(Base.Pay,Payments.Over.Base.Pay,Average.Benefit.Cost,Overtime.Pay))
cmatrix <- round(cor(datasub),4)
cmatrix
cmatrix_melted <- melt(cmatrix)
par(las=2)
ggplot(cmatrix_melted, aes(x=Var1, y=Var2, fill=value)) + geom_tile() #Building a heatmap


```

#### Description:
* * *
This heatmap represents the correlation between the crucial factors that can provide fo an individual's living costs in the city. Some forms of payment such as avg. benefit cost are slightly correlated with base pay ndicating both values are somewhat proportional.


#### 6. To derive a relationship between final quarterly payment and base pay of a particular department and predict the base pay of its employees.
```{r}
library(dplyr)
library(magrittr)
library(ggplot2)
library(caTools)
library(reshape2)
data <- read.csv("data_clean.csv")
pd_2015 <- data %>% select(Q1.Payments:Q4.Payments,Projected.Annual.Salary, Base.Pay) %>% filter(data$Year=='2015', data$Department.Title=='police (lapd)')
corr_df <- cor(pd_2015)
corr_df

#Using the corr_df matrix we can observe the strong correlation between base pay and Q4 payment

#building model
set.seed(123)
split = sample.split(pd_2015$Base.Pay, SplitRatio = 0.90)
training_set = subset(pd_2015, split == TRUE)
test_set = subset(pd_2015, split == FALSE)

#Fitting Simple Linear Regression to the Training set
regressor = lm(formula = Base.Pay ~  Q4.Payments, data = training_set)
summary(regressor)

#Predicting the Test set results
y_pred = predict(regressor, newdata = test_set)
y_pred

options("scipen"=100, "digits"=4)

g1<-ggplot() +
  geom_point(aes(x = training_set$Q4.Payments, y = training_set$Base.Pay),
             colour = 'blue') +
  geom_line(aes(x = training_set$Q4.Payments, y = predict(regressor, newdata = training_set)),
            colour = 'red') +
  ggtitle('Base Pay v/s Q4 Payments - Training') +
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) +
  xlab('Final Quaterly Payments') +
  ylab('Base Pay (in USD)')

# Test set
g2<-ggplot() +
  geom_point(aes(x = test_set$Q4.Payments, y = test_set$Base.Pay),
             colour = 'green') +
  geom_line(aes(x = training_set$Q4.Payments, y = predict(regressor, newdata = training_set)),
            colour = 'red') +
  ggtitle('Base Pay v/s Q4 Payments - Test') +
  xlab('Final Quaterly Payments') +
  ylab('Base Pay (in USD)')

par(mfrow=c(2,2))
g1
g2
```

#### Description:
* * *
From the correlation matrix we can observe a strong correlation between the factor - Base pay and Q4 Payment. Employees recieve almost 1/4th of their total promised pay during the final quarter even inclusive of their overtime or extra pay. This indicates a strong linear relationship i.e. as Q4 payments increase, base pay also increases.
Hence we have used this factor to predict the base pay of some of the employees belonging to a particular department(police, 2015 in this case). Using simple linear regression and with Q4.Payments as predictor variable we predict the Base.Pay as outcome. The predicted values are indicated by the points in green.


## Conclusion:
* * * 
In this analysis, we have been able to identify the job titles in LA offering salary less than the minimum cost of living. We were also able to identify the departments that have unusual or large differences in their salaries i.e. wage gap and compare it with other departments. Finally we were able to predict the base salary of some of the employees using their final quarterly payment.


